---
title: "Statistical Methods for Linguistic Data"
subtitle: "CLA 2026 Crash Course Practical Component"
author: "Lindsay Hracs"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: false
    highlight: kate
    code_folding: show
---

```{css, echo=FALSE}
/* Nord-inspired styling */
body { font-family: 'Segoe UI', Calibri, sans-serif; color: #2E3440; }
h1, h2, h3 { color: #2E3440; }
h2 { border-bottom: 3px solid #88C0D0; padding-bottom: 6px; }
h3 { color: #4C566A; }
.banner {
  background-color: #3B4252;
  border-left: 6px solid #88C0D0;
  padding: 16px 20px;
  border-radius: 6px;
  color: #ECEFF4;
  margin-bottom: 20px;
}
.note {
  background-color: #E5E9F0;
  border-left: 4px solid #88C0D0;
  padding: 12px 16px;
  border-radius: 4px;
  margin: 12px 0;
}
```

<div class="banner">

### Local backup version

This RMarkdown file is a local backup of the Colab notebook for use without internet access.
All datasets are loaded from local CSV files; see **Section 0** for setup instructions.

</div>

---

## About this document

This notebook accompanies the workshop lecture. We will work through five analyses using
synthetic data from a synthetic study on the focus sensitive particle (FSP) *only* in English.

---

### Synthetic study 1: Experimental study

**Research question:** What effects do we observe in the different structural positions of the
focus sensitive particle *only*?

**Participants:** 83 adult native English speakers recruited from a university population.

**Design:** 2 x 4 factorial

- **Context** (2 levels): *match* (sentence is TRUE given context) vs. *mismatch* (sentence is FALSE given context)
- **FSP position** (4 levels): the position of *only* in the sentence
  - `FSP-S-V-O`: *Only Dale drinks coffee.*
  - `S-FSP-V-O`: *Dale only drinks coffee.*
  - `S-V-FSP-O`: *Dale drinks only coffee.*
  - `S-V-O-FSP`: *Dale drinks coffee only.*

| Task | Measure | Outcome variable | Type |
|------|---------|-----------------|------|
| Sentence production | Peak F0 on focused constituent (Hz) | `peak_f0` | Continuous |
| Acceptability judgment | 1-7 Likert rating | `rating` | Ordinal |
| Truth Value Judgment Task (TVJT) | Accuracy (correct/incorrect) | `correct` | Binary |
| TVJT | Response time (ms) | `rt_ms` | Continuous |

### Synthetic study 2: Corpus study

**Research question:** Does the frequency of FSP position types vary across registers?

| Corpus type | Register | Example corpora |
|-------------|----------|-----------------|
| Newspaper | Written formal | The National Chronicle, The Daily Register |
| Social media written | Written informal | RedditLing, BlueSky Academic |
| Legal proceedings | Spoken formal | Federal Court Transcripts, Provincial Court Transcripts |
| Social media spoken | Spoken informal | LinguistTok, GrammarReels |

**Outcome variable:** `count` (non-negative integer); overdispersed, requiring negative binomial regression.

---

## 0. Setup

### Install and load packages

```{r setup, message=FALSE, warning=FALSE}
if (!require("pacman")) install.packages("pacman")

pacman::p_load(
  lme4,      # linear and generalized linear mixed effects models
  lmerTest,  # p-values and degrees of freedom for lmer()
  ggplot2,   # plotting
  dplyr,     # data wrangling
  MASS,      # negative binomial
  ordinal,   # cumulative link models (ordinal regression)
  patchwork, # combining multiple plots
  viridis    # visually-impaired-friendly colour palette
)

cat("All packages successfully loaded!\n")
```

### Load local data files

<div class="note">
**Local use:** Download the five CSV files from the workshop GitHub repository and place them
in the same folder as this RMD file before knitting, or update the file paths below.

GitHub: https://github.com/lhracs/CLA2026_crash_course
</div>

```{r load-data, message=FALSE}
participants  <- read.csv("participants_synth.csv",  stringsAsFactors = FALSE)
production    <- read.csv("production_synth.csv",    stringsAsFactors = FALSE)
acceptability <- read.csv("acceptability_synth.csv", stringsAsFactors = FALSE)
tvjt          <- read.csv("tvjt_synth.csv",          stringsAsFactors = FALSE)
corpus        <- read.csv("corpus_synth.csv",        stringsAsFactors = FALSE)

cat("All datasets successfully loaded!\n")
cat("  participants: ", nrow(participants), "rows\n")
cat("  production:  ", nrow(production),   "rows\n")
cat("  acceptability:", nrow(acceptability),"rows\n")
cat("  tvjt:        ", nrow(tvjt),         "rows\n")
cat("  corpus:      ", nrow(corpus),       "rows\n")
```

---

## 1. Inspect and format the data

```{r inspect, results='hold'}
head(participants)
str(participants)

head(production)
str(production)

head(acceptability)
str(acceptability)

head(tvjt)
str(tvjt)
```

<div class="note">
**Key question before modelling:** What type is each outcome variable?

- `peak_f0`: continuous (Hz)
- `rating`: ordinal (1-7 Likert scale, NOT continuous)
- `correct`: binary (0 or 1)
- `rt_ms`: continuous but right-skewed (reaction times)

Each type requires a different model family. This is the first decision in the analysis pipeline.
</div>

```{r factor-levels}
# Reference: FSP-S-V-O (focus sensitive particle directly precedes subject)
production$fsp_position    <- factor(production$fsp_position,
                                     levels = c("FSP-S-V-O","S-FSP-V-O","S-V-FSP-O","S-V-O-FSP"))
acceptability$fsp_position <- factor(acceptability$fsp_position,
                                     levels = c("FSP-S-V-O","S-FSP-V-O","S-V-FSP-O","S-V-O-FSP"))
tvjt$fsp_position          <- factor(tvjt$fsp_position,
                                     levels = c("FSP-S-V-O","S-FSP-V-O","S-V-FSP-O","S-V-O-FSP"))

# Context reference level: mismatch
production$context    <- factor(production$context,    levels = c("mismatch","match"))
acceptability$context <- factor(acceptability$context, levels = c("mismatch","match"))
tvjt$context          <- factor(tvjt$context,          levels = c("mismatch","match"))

# Rating must be an ordered factor for ordinal regression
acceptability$rating <- factor(acceptability$rating, levels = 1:7, ordered = TRUE)

cat("Factor levels set!\n")
```

---

## 2. Visualize before you model

Always plot your data before fitting any model. Summary statistics can be deeply misleading.

### 2.1 Distribution of RT: Histogram

```{r hist-rt, fig.width=8, fig.height=4}
ggplot(tvjt, aes(x = rt_ms)) +
  geom_histogram(bins = 40, fill = viridis(1, option = "D"), color = "#2E3440", alpha = 0.85) +
  geom_vline(aes(xintercept = mean(rt_ms)), color = viridis(1, option = "A"),
             linewidth = 1, linetype = "dashed") +
  labs(
    title    = "Distribution of response times",
    subtitle = "Dashed line = mean. Notice the right skew and outliers.",
    x        = "RT (ms)",
    y        = "Count"
  ) +
  theme_minimal(base_size = 13)
```

<div class="note">
**What to notice:**

- The distribution is right-skewed; a long tail of slow responses
- The mean (dashed line) is pulled toward the tail
- A few extreme outliers are visible on the right

This tells us we should log-transform RT before fitting a linear model.
</div>

### 2.2 Ratings by FSP position and context: Violin + boxplot

```{r violin-ratings, fig.width=9, fig.height=5}
ggplot(acceptability, aes(x = fsp_position, y = as.numeric(rating), fill = context)) +
  geom_violin(position = position_dodge(0.8), alpha = 0.7) +
  geom_boxplot(position = position_dodge(0.8), width = 0.2,
               outlier.shape = 21, outlier.size = 1.5) +
  scale_fill_viridis_d(option = "D", begin = 0.3, end = 0.75) +
  labs(
    title    = "Acceptability ratings by FSP position and context",
    subtitle = "1 = very unacceptable, 7 = very acceptable",
    x        = "FSP position",
    y        = "Rating (1-7)",
    fill     = "Context"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

<div class="note">
**What to notice:**

- Match contexts are rated higher than mismatch contexts across all FSP positions
- `S-FSP-V-O` (preverbal *only*) shows the most spread and the lowest match ratings
- Ratings cluster at the high end in match conditions; ceiling effects are visible
- This is why we need ordinal regression: the distributions are not normal, and the gaps between levels are not equal
</div>

### 2.3 F0 by FSP position: Strip chart

```{r strip-f0, fig.width=8, fig.height=5}
ggplot(production, aes(x = fsp_position, y = peak_f0, colour = fsp_position)) +
  geom_jitter(width = 0.2, alpha = 0.35, size = 1.5) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.5,
               colour = "#2E3440", linewidth = 0.8) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.25,
               colour = "#2E3440", linewidth = 0.8) +
  scale_colour_viridis_d(option = "D") +
  labs(
    title    = "Peak F0 on focused constituent by FSP position",
    subtitle = "Crossbar = mean +/- SE. Individual observations shown as points.",
    x        = "FSP position",
    y        = "Peak F0 (Hz)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

<div class="note">
**What to notice:**

- There is a clear gradient: `FSP-S-V-O` shows the highest F0, `S-V-O-FSP` the lowest
- Individual variation is large; some participants have much higher baseline F0
- A few outlier points are visible at the extremes
- The group pattern is consistent despite individual differences; this is what random effects model
</div>

### 2.4 Accuracy by FSP position and context: Bar chart

```{r bar-acc, fig.width=9, fig.height=5}
tvjt_summary <- tvjt %>%
  group_by(fsp_position, context) %>%
  summarise(mean_acc = mean(correct), se = sd(correct) / sqrt(n()), .groups = "drop")

ggplot(tvjt_summary, aes(x = fsp_position, y = mean_acc, fill = context)) +
  geom_col(position = position_dodge(0.7), width = 0.6) +
  geom_errorbar(aes(ymin = mean_acc - se, ymax = mean_acc + se),
                position = position_dodge(0.7), width = 0.25) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "#4C566A") +
  scale_fill_viridis_d(option = "D", begin = 0.3, end = 0.75) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  labs(
    title    = "TVJT accuracy by FSP position and context",
    subtitle = "Dashed line = chance (50%). Error bars = +/- 1 SE.",
    x        = "FSP position",
    y        = "Mean accuracy",
    fill     = "Context"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

<div class="note">
**What to notice:**

- All conditions are well above chance; participants can assign truth values to sentences with *only*
- Accuracy drops for mismatch conditions, especially `S-V-FSP-O`
- We will model this with logistic regression in Section 5
</div>

---

## 3. Linear mixed effects model: Peak F0

**Outcome:** `peak_f0` (continuous, Hz)  
**Question:** Does FSP position predict where pitch accent lands?

```{r model-f0}
model_f0 <- lmer(peak_f0 ~ fsp_position + context + (1 | participant_id) + (1 | item_id),
                 data = production,
                 REML = TRUE)

summary(model_f0)
```

```{r ci-f0}
confint(model_f0, method = "Wald")
```

<div class="note">
**A note on NA sigma values:** Standard deviations cannot be negative. Wald intervals assume
estimates could fall anywhere on a number line and rather than return intervals that include
impossible values, R returns `NA`. The fixed effects intervals are unaffected. Use
`confint(model, method = "profile")` if you need intervals for variance components.
</div>

```{r diag-f0, fig.width=9, fig.height=4}
par(mfrow = c(1, 2))
plot(model_f0,    main = "Residuals vs. Fitted: F0 model")
qqnorm(resid(model_f0), main = "Q-Q plot: F0 model residuals")
qqline(resid(model_f0), col = viridis(1, option = "A"))
par(mfrow = c(1, 1))
```

```{r cohens-d-f0}
residual_sd <- sigma(model_f0)
coefs       <- fixef(model_f0)

cat("Cohen's d for fixed effects:\n\n")
for (nm in names(coefs)) {
  d <- coefs[nm] / residual_sd
  cat(sprintf("  %-30s d = %6.3f\n", nm, d))
}
cat("\nd <= 0.2 = small | d = 0.5 = medium | d >= 0.8 = large\n")
```

---

## 4. Ordinal regression: Acceptability ratings

**Outcome:** `rating` (ordered, 1-7)  
**Question:** Does FSP position and context affect acceptability?

### Why not lmer()?

```{r wrong-model, warning=FALSE}
acceptability$rating_num <- as.numeric(acceptability$rating)

incorrect_mod <- lmer(rating_num ~ fsp_position + context + (1 | participant_id) + (1 | item_id),
                      data = acceptability, REML = TRUE)
```

```{r qq-wrong, fig.width=5, fig.height=5}
qqnorm(resid(incorrect_mod), main = "Q-Q plot for incorrect model (lmer on Likert ratings)")
qqline(resid(incorrect_mod), col = viridis(1, option = "A"))
```

<div class="note">
**Q-Q plot interpretation:** Points should fall on the diagonal if residuals are normally
distributed. The deviation at both ends indicates the residuals are not normal, which is expected
when a bounded ordinal outcome is forced into a linear model.
</div>

```{r impossible-values}
range(fitted(incorrect_mod))
```

<div class="note">
**Impossible predictions:** The range above shows fitted values outside the 1-7 scale. A model
that predicts impossible values is misspecified regardless of whether the p-values look significant.
</div>

```{r dist-ratings, fig.width=10, fig.height=6}
ggplot(acceptability, aes(x = rating)) +
  geom_bar(fill = viridis(1, option = "D")) +
  facet_wrap(~ fsp_position + context) +
  labs(title = "Distribution of ratings by condition") +
  theme_minimal(base_size = 12)
```

<div class="note">
**Distribution of ratings:** Ratings cluster at discrete values and do not spread smoothly
across the scale the way a continuous variable should. Ordinal regression models these
discrete categories directly.
</div>

```{r means-demo, fig.width=7, fig.height=4}
set.seed(8675309)
demo <- data.frame(
  condition = rep(c("A", "B"), each = 50),
  rating    = c(
    sample(c(1, 7), 50, replace = TRUE),
    sample(3:5,     50, replace = TRUE)
  )
)

demo %>% group_by(condition) %>% summarise(mean = mean(rating))

ggplot(demo, aes(x = factor(rating), fill = condition)) +
  geom_bar(position = "dodge") +
  scale_fill_viridis_d(option = "D", begin = 0.3, end = 0.75) +
  labs(title    = "Same mean, completely different distributions",
       subtitle = "Linear regression cannot distinguish these -- ordinal regression can",
       x = "Rating", y = "Count", fill = "Condition") +
  theme_minimal(base_size = 13)
```

<div class="note">
**The means problem:** Conditions A and B have identical means so a linear model would find no
difference. But their distributions are completely different. Ordinal regression can detect this.
</div>

### Fit the correct model

```{r model-rating}
model_rating <- clmm(rating ~ fsp_position + context + (1 | participant_id) + (1 | item_id),
                     data = acceptability,
                     link = "logit")

summary(model_rating)
```

```{r lrt-rating}
model_rating_null <- clmm(rating ~ context + (1 | participant_id) + (1 | item_id),
                           data = acceptability, link = "logit")

anova(model_rating_null, model_rating)
```

<div class="note">
**Interpreting the likelihood ratio test:** Adding FSP position significantly improves model fit.
This tells us FSP position explains variance in acceptability over and above context alone.
</div>

---

## 5. Logistic mixed effects model: TVJT accuracy

**Outcome:** `correct` (binary: 0 or 1)  
**Question:** Does FSP position and context predict whether participants correctly judge sentences?

```{r model-acc, warning=TRUE}
model_acc <- glmer(correct ~ fsp_position + context + (1 | participant_id) + (1 | item_id),
                   data = tvjt,
                   family = binomial(link = "logit"))

summary(model_acc)
```

<div class="note">
**Singular fit warning:** The participant random effect variance is essentially zero. The model
is telling you there is no meaningful variation between participants in baseline accuracy. Refit
without the participant random intercept and compare with AIC.
</div>

```{r simplify-acc}
model_acc_nopart <- glmer(correct ~ fsp_position + context + (1 | item_id),
                          data = tvjt, family = binomial)

AIC(model_acc, model_acc_nopart)
```

<div class="note">
The simpler model has a lower AIC so go with `model_acc_nopart`. The singular fit warning is
resolved, the model is more parsimonious, and AIC supports the simpler structure.
</div>

```{r summary-acc-nopart}
summary(model_acc_nopart)
```

```{r plogis-acc}
intercept <- fixef(model_acc_nopart)["(Intercept)"]

# Reference condition
cat("P(correct): FSP-S-V-O, mismatch = ", round(plogis(intercept), 3), "\n")
cat("P(correct): FSP-S-V-O, match =    ", round(plogis(intercept +
                                                   fixef(model_acc_nopart)["contextmatch"]), 3), "\n\n")

# FSP position effects -- only S-V-FSP-O is significant (p < .001)
cat("P(correct): S-FSP-V-O, mismatch = ", round(plogis(intercept +
                                                   fixef(model_acc_nopart)["fsp_positionS-FSP-V-O"]), 3), "\n")
cat("P(correct): S-FSP-V-O, match =    ", round(plogis(intercept +
                                                   fixef(model_acc_nopart)["fsp_positionS-FSP-V-O"] +
                                                   fixef(model_acc_nopart)["contextmatch"]), 3), "\n\n")

cat("P(correct): S-V-FSP-O, mismatch = ", round(plogis(intercept +
                                                   fixef(model_acc_nopart)["fsp_positionS-V-FSP-O"]), 3), "\n")
cat("P(correct): S-V-FSP-O, match =    ", round(plogis(intercept +
                                                   fixef(model_acc_nopart)["fsp_positionS-V-FSP-O"] +
                                                   fixef(model_acc_nopart)["contextmatch"]), 3), "\n\n")

cat("P(correct): S-V-O-FSP, mismatch = ", round(plogis(intercept +
                                                   fixef(model_acc_nopart)["fsp_positionS-V-O-FSP"]), 3), "\n")
cat("P(correct): S-V-O-FSP, match =    ", round(plogis(intercept +
                                                   fixef(model_acc_nopart)["fsp_positionS-V-O-FSP"] +
                                                   fixef(model_acc_nopart)["contextmatch"]), 3), "\n")
```

<div class="note">
**Key takeaway:** Converting from log-odds to probabilities makes the results interpretable.
The S-V-FSP-O mismatch condition has the lowest accuracy, meaning participants have a harder
time correctly judging sentences with *only* directly preceding the object as FALSE.
</div>

---

## 6. Linear mixed effects model: Response time

**Outcome:** `rt_ms` (continuous, right-skewed)  
**Question:** Does FSP position and context predict response time?

```{r compare-rt, fig.width=10, fig.height=4}
p1 <- ggplot(tvjt, aes(x = rt_ms)) +
  geom_histogram(bins = 40, fill = viridis(1, option = "D"), colour = "#2E3440", alpha = 0.85) +
  labs(title = "Raw RT (ms)", x = "RT (ms)", y = "Count") +
  theme_minimal(base_size = 12)

p2 <- ggplot(tvjt, aes(x = log(rt_ms))) +
  geom_histogram(bins = 40, fill = viridis(1, option = "E"), colour = "#2E3440", alpha = 0.85) +
  labs(title = "Log-transformed RT", x = "log(RT)", y = "Count") +
  theme_minimal(base_size = 12)

p1 + p2
```

<div class="note">
**Notice:** Log-transformation makes the distribution much more symmetric and closer to normal.
This is what the linear model assumes for its residuals.
</div>

```{r model-rt}
tvjt$log_rt <- log(tvjt$rt_ms)

model_rt <- lmer(log_rt ~ fsp_position + context + (1 | participant_id) + (1 | item_id),
                 data = tvjt,
                 REML = TRUE)

summary(model_rt)
```

```{r back-transform-rt}
cat("Multiplicative RT effects (relative to reference condition):\n\n")

fe <- fixef(model_rt)
for (nm in names(fe)) {
  cat(sprintf("  %-35s exp(b) = %.3f\n", nm, exp(fe[nm])))
}
```

<div class="note">
**Interpreting multiplicative effects:**

- exp(b) = 1.0 means no effect on RT
- exp(b) = 1.1 means RT is 10% longer
- exp(b) = 0.9 means RT is 10% shorter
</div>

```{r diag-rt, fig.width=9, fig.height=4}
par(mfrow = c(1, 2))
plot(model_rt,    main = "Residuals vs. Fitted: RT model",
     col = "grey40", col.smooth = "transparent")
qqnorm(resid(model_rt), main = "Q-Q plot: RT model residuals")
qqline(resid(model_rt), col = viridis(1, option = "A"))
par(mfrow = c(1, 1))
```

---

## 7. Count models: FSP occurrence in corpus data

**Outcome:** `count` (non-negative integer)  
**Question:** Does FSP position frequency vary across corpus types (register)?

### 7.1 Inspect the data

```{r corpus-inspect}
corpus$corpus_type  <- factor(corpus$corpus_type,
                               levels = c("newspaper","social_media_written",
                                          "legal","social_media_spoken"))
corpus$fsp_position <- factor(corpus$fsp_position,
                               levels = c("FSP-S-V-O","S-FSP-V-O","S-V-FSP-O","S-V-O-FSP"))

cat("Loaded corpus data:", nrow(corpus), "rows\n")

corpus %>%
  group_by(corpus_type) %>%
  dplyr::slice(1) %>%
  dplyr::select(corpus_type, corpus, word_count, excerpt)
```

### 7.2 Visualize

```{r corpus-plot, fig.width=10, fig.height=5}
corpus_summary <- corpus %>%
  group_by(corpus_type, fsp_position) %>%
  summarise(mean_count = mean(count), se = sd(count) / sqrt(n()), .groups = "drop")

ggplot(corpus_summary, aes(x = corpus_type, y = mean_count, fill = fsp_position)) +
  geom_col(position = position_dodge(0.75), width = 0.7) +
  geom_errorbar(aes(ymin = mean_count - se, ymax = mean_count + se),
                position = position_dodge(0.75), width = 0.25) +
  scale_fill_viridis_d(option = "D") +
  labs(
    title    = "Mean FSP occurrence count per document by corpus type",
    subtitle = "Error bars = +/- 1 SE. Note: document length varies by corpus type.",
    x        = "Corpus type",
    y        = "Mean count per document",
    fill     = "FSP position"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 15, hjust = 1))
```

### 7.3 Check for overdispersion

```{r dispersion-check}
dispersion_check <- corpus %>%
  group_by(fsp_position) %>%
  summarise(
    mean_count = mean(count),
    var_count  = var(count),
    ratio      = var(count) / mean(count),
    .groups    = "drop"
  )

print(dispersion_check)
```

<div class="note">
**What to notice:** The variance/mean ratio should be approximately 1 for Poisson data.
Ratios well above 1 indicate overdispersion. We need negative binomial regression.
</div>

### 7.4 Fit Poisson and check

```{r model-poisson, warning=FALSE}
model_poisson <- glmer(count ~ fsp_position + corpus_type + offset(log(word_count)) +
                         (1 | corpus),
                       data   = corpus,
                       family = poisson(link = "log"))

cat("Residual deviance:", deviance(model_poisson), "\n")
cat("Residual degrees of freedom:", df.residual(model_poisson), "\n")
cat("Ratio (should be ~1 for Poisson):", round(deviance(model_poisson) / df.residual(model_poisson), 2), "\n")
```

<div class="note">
If the ratio is much greater than 1, the Poisson model is a poor fit. Use negative binomial.
</div>

### 7.5 Fit negative binomial

```{r model-nb}
model_nb <- glmer.nb(count ~ fsp_position + corpus_type + offset(log(word_count)) +
                       (1 | corpus),
                     data = corpus)

summary(model_nb)
```

```{r rate-ratios}
cat("Rate ratios (exp(b)) relative to reference (FSP-S-V-O, newspaper):\n\n")
fe <- fixef(model_nb)
for (nm in names(fe)) {
  cat(sprintf("  %-45s exp(b) = %.3f\n", nm, exp(fe[nm])))
}
```

<div class="note">
**Interpreting rate ratios:**

- exp(b) = 1.0 means no difference in rate
- exp(b) = 2.0 means twice the rate of the reference condition
- exp(b) = 0.5 means half the rate of the reference condition
</div>

```{r aic-compare}
cat("Poisson AIC:          ", AIC(model_poisson), "\n")
cat("Negative binomial AIC:", AIC(model_nb), "\n")
cat("\nLower AIC = better fit. A difference > 2 is meaningful.\n")
```

---

## Summary

| Task | Outcome | Model | R function |
|------|---------|-------|------------|
| Production | Peak F0 (Hz) | Linear mixed effects | `lmer()` |
| Acceptability | Rating (1-7) | Ordinal regression | `clmm()` |
| TVJT | Accuracy (0/1) | Logistic mixed effects | `glmer()` |
| TVJT | Response time (ms) | Linear mixed effects on log RT | `lmer()` |
| Corpus | FSP occurrence count | Negative binomial mixed effects | `glmer.nb()` |

### Key methodological takeaways

**Your outcome variable drives everything**

- Every modelling decision started with asking what type of variable your outcome is
- Getting this wrong may produce results that look fine, but in fact, are not

**Always visualize before you model**

- The RT histogram revealed right skew before we fitted anything, which is why we log-transformed
- Rating distributions showed discrete clustering, which is why we used `clmm()` not `lmer()`
- Plots can make the modelling decisions obvious before running a single model

**It is important to consider random effects**

- Models included random intercepts for the appropriate grouping factors
- The logistic model flagged a singular fit, which led us to simplify and produce a better model

**Model comparison is part of the analysis**

- Likelihood ratio test to justify the ordinal model over the null
- AIC to confirm choice between Poisson and negative binomial
- AIC to resolve the singular fit in the logistic model
- Model selection is built into the pipeline, not an afterthought

**Multiple tasks reveal a richer picture**

- FSP position affected F0, acceptability, accuracy, and RT, but not always the same positions
- S-V-FSP-O was the hardest condition in accuracy and RT
- S-FSP-V-O showed an acceptability penalty but not an accuracy penalty
- A single task and a single model would have missed these nuances

**Assumption checking leads to better decisions**

- Residuals and Q-Q plots for linear models
- Dispersion test for count data
- Singular fit check for all mixed models

---

*Workshop materials by Lindsay Hracs for CLA 2026 Crash Course*  
*Contact: lindsay[dot]hracs[at]ucalgary[dot]ca*
